### 2015-05-11 doc/thesis/introduction.tex
Renaming thesis directory

Dynamically typed languages such as Lua avoid static types in favor of
simplicity and flexibility, because the absence of static types means
that programmers do not need to bother with abstracting types that
should be validated by a type checker.
Instead, dynamically typed languages use run-time \emph{type tags}
to classify the values they compute, so their implementation can use
these tags to perform run-time (or dynamic) type checking
\citep{pierce2002tpl}.

@@

Thus, early error detection and better program structure are two
advantages of static type checking that can lead programmers to
migrate their code from a dynamically typed to a statically
typed language, when their simple scripts evolve into complex programs
\citep{tobin-hochstadt2006ims}.
Dynamically typed languages certainly help programmers during the
beginning of a project, because their simplicity and flexibility
allows quick development and makes it easier to change code according to
changing requirements.
However, programmers tend to migrate from dynamically typed to
statically typed code as soon as the project has consolidated its
requirements, because the robustness of static types helps
programmers link requirements to abstractions.
This migration usually involves different languages that have
different syntaxes and semantics, which usually requires a complete
rewrite of existing programs instead of incremental evolution from
dynamic to static types.

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing:
\emph{optional type systems} \citep{bracha2004pluggable} and
\emph{gradual typing} \citep{siek2006gradual} are two similar
approaches for blending static and dynamic typing in the same
language.
The aim of both approaches is to offer programmers the option
to use type annotations where static typing is needed,
allowing the incremental migration from dynamic to static typing.
The difference between these two approaches is the way they treat
run-time semantics.
While optional type systems do not affect run-time semantics,
gradual typing uses run-time checks to ensure that dynamically typed
code does not violate the invariants of statically typed code.

Programmers and researchers sometimes use the term \emph{gradual typing}
to mean the incremental evolution of dynamically typed code into
statically typed code.
For this reason, gradual typing may also refer to optional type
systems and other approaches that blend static and dynamic typing to
help programmers incrementally migrate from dynamic to static typing
without having to switch to a different language, though all these
approaches differ in the way they handle static and dynamic typing
together.
We use the term \emph{gradual typing} to refer to the work of
\citet{siek2006gradual}.

@@

Typed Lua performs a very limited form of local type inference
\citep{pierce2000lti}, as static typing does not necessarily mean
that programmers need to insert type annotations in the code.
Several statically typed languages such as Haskell provide some
amount of type inference that automatically deduces the types of
expressions.
Still, Typed Lua only requires a small amount of type annotations
due to the nature of its optional type system.

>>>>

Dynamically typed languages such as Lua avoid static types in favor of
simplicity and flexibility, because the absence of static types means
that programmers do not need to bother with abstracting types that
should be validated by a type checker.
Instead, dynamically typed languages use run-time \emph{type tags}
to classify the values they compute, so their implementation can use
these tags to perform run-time (or dynamic) type checking
\cite{pierce2002tpl}.

@@

Thus, early error detection and better program structure are two
advantages of static type checking that can lead programmers to
migrate their code from a dynamically typed to a statically
typed language, when their simple scripts evolve into complex programs
\cite{tobin-hochstadt2006ims}.
Dynamically typed languages certainly help programmers during the
beginning of a project, because their simplicity and flexibility
allows quick development and makes it easier to change code according to
changing requirements.
However, programmers tend to migrate from dynamically typed to
statically typed code as soon as the project has consolidated its
requirements, because the robustness of static types helps
programmers link requirements to abstractions.
This migration usually involves different languages that have
different syntaxes and semantics, which usually requires a complete
rewrite of existing programs instead of incremental evolution from
dynamic to static types.

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing:
\emph{optional type systems} \cite{bracha2004pluggable} and
\emph{gradual typing} \cite{siek2006gradual} are two similar
approaches for blending static and dynamic typing in the same
language.
The aim of both approaches is to offer programmers the option
to use type annotations where static typing is needed,
allowing the incremental migration from dynamic to static typing.
The difference between these two approaches is the way they treat
run-time semantics.
While optional type systems do not affect run-time semantics,
gradual typing uses run-time checks to ensure that dynamically typed
code does not violate the invariants of statically typed code.

Programmers and researchers sometimes use the term \emph{gradual typing}
to mean the incremental evolution of dynamically typed code into
statically typed code.
For this reason, gradual typing may also refer to optional type
systems and other approaches that blend static and dynamic typing to
help programmers incrementally migrate from dynamic to static typing
without having to switch to a different language, though all these
approaches differ in the way they handle static and dynamic typing
together.
We use the term \emph{gradual typing} to refer to the work of
Siek and Taha \cite{siek2006gradual}.

@@

Typed Lua performs a very limited form of local type inference
\cite{pierce2000lti}, as static typing does not necessarily mean
that programmers need to insert type annotations in the code.
Several statically typed languages such as Haskell provide some
amount of type inference that automatically deduces the types of
expressions.
Still, Typed Lua only requires a small amount of type annotations
due to the nature of its optional type system.

### 2015-05-11 doc/thesis/system.tex
Final touches before sending to juries
Renaming thesis directory

After type checking the right-hand side, our type system type checks the list
of left-hand values.
In our example, the left-hand side of the assignment has type 
$\Integer \times \String \times \Value{*}$.
Note that our type system uses the type $\Value{*}$ to discard extra values.
The rule that type checks lists of left-hand values introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
Our example uses the rule \textsc{T-LHSLIST} to type
checks a list of left-hand values:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env_{1}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

After type checking the right-hand side and the left-hand side of an assignment,
our type system checks whether their types are consistent.
The rule \textsc{T-ASSIGNMENT} is the general rule that express this idea:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
\end{array}
\]

@@

The simplest expression involving tables in the empty table constructor.
Its type checking rule is straightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env_{1}, \penv \vdash \{\}:\{\}_{unique}, \env_{1}
\end{array}
\]

@@

The function \emph{read} uses the auxiliary function \emph{proj}
to project a union of first-level types, based on an union of
second-level types and an index from a projection type.
In Section \ref{sec:fap} we will discuss how our type system uses
projection types to handle overloaded return types.
We can define \emph{proj} as follows:
\begin{align*}
proj(T_{1} \times ... \times T_{n}{*}, i) & = \left\{
\begin{array}{ll}
T_{i} & \text{if $i < n$}\\
T_{i} \cup \Nil & \text{if $i = n$}
\end{array} \right.
\\
proj(S_{1} \sqcup S_{2}, i) & = proj(S_{1}, i) \cup proj(S_{2}, i)
\end{align*}

>>>>

After type checking the right-hand side, our type system type checks the list
of left-hand values.
In our example, the left-hand side of the assignment has type 
$\Integer \times \String \times \Value{*}$.
Note that our type system uses the type $\Value{*}$ to discard extra values.
The rule that type checks lists of left-hand values introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
Our example uses the rule \textsc{T-LHSLIST} to type
check a list of left-hand values:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env_{1}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

After type checking the right-hand side and the left-hand side of an assignment,
our type system checks whether their types are consistent.
The rule \textsc{T-ASSIGNMENT} is the general rule that expresses this idea:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
\end{array}
\]

@@

The simplest expression involving tables is the empty table constructor.
Its type checking rule is straightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env_{1}, \penv \vdash \{\}:\{\}_{unique}, \env_{1}
\end{array}
\]

@@

The function \emph{read} uses the auxiliary function \emph{proj}
to project a union of first-level types, based on an union of
second-level types and an index from a projection type.
In Section \ref{sec:fap} we will discuss how our type system uses
projection types to handle overloaded return types.
We can define \emph{proj} as follows:
\begin{align*}
proj(T_{1} \times ... \times T_{n} \times T{*}, i) & =  T_{i} \;\;\; \text{if $i <= n$}\\
proj(S_{1} \sqcup S_{2}, i) & = proj(S_{1}, i) \cup proj(S_{2}, i)
\end{align*}

### 2015-05-11 doc/thesis/typedlua.tex
Renaming thesis directory

Typed Lua is an optional type system for Lua, and its main goal is to
provide static type checking for Lua.
To do that, Typed Lua extends the syntax of Lua 5.3 to introduce
optional type annotations, and performs local type inference \citep{pierce2000lti}
to detect more precise types for unannotated expressions.
Even though the compiler warns the programmer about type errors,
it always removes the type annotations to generate Lua code that
runs in unmodified Lua implementations.

Another goal of Typed Lua is to be backwards compatible with Lua.
This means that any Lua code is valid Typed Lua code.
To be backwards compatible with Lua, the syntactic extensions introduced
by Typed Lua does not include new reserved words.
Appendix \ref{app:syntax} presents the complete syntax of Typed Lua
in extended BNF.

We use the consistent-subtyping relation of gradual typing
\citep{siek2007objects,siek2013mutable} to formalize Typed Lua,
though it does not insert run-time checks in the gradual typing style.
In gradual typing, run-time checks inspect the interaction between
dynamically typed and statically typed code to guarantee that dynamically
typed code does not violate statically typed code during run-time.
We did not insert run-time checks at this moment because they can decrease
run-time performance \citep{allende2013cis}.
We believe that a careful evaluation of run-time checks should be done before
inserting them in the type system.
However, this evaluation is out of scope of this work.

Unlike Dart \citep{dart} and TypeScript \citep{typescript},
we are designing Typed Lua aiming soundness to make it possible to
switch Typed Lua from optional typing to gradual typing in the future.
A sound type system is a prerequisite to insert run-time checks after
static type checking, because a sound type system ensures that
statically typed code will not throw type errors during run-time.

@@

Lua programmers often overload the input parameters of functions,
and use the \texttt{type} function to inspect the tag of the
input parameters to take different actions depending on what
those tags are.
The simplest case overloads on just a single parameter:
\begin{verbatim}
    local function overload (s1:string, s2:string|integer)
      if type(s2) == "string" then
        return s1 .. s2
      else
        -- string.rep : (string, integer, string) -> (string)
        return string.rep(s1, s2)
      end
    end
\end{verbatim}

Typed Lua has a small set of \texttt{type} predicates that allows
programmers to constrain the type of a local variable inside a condition.
This example uses the predicate \texttt{type(v) == "string"} that
constrains the type of \texttt{v} from \texttt{string|t} to
\texttt{string} when the predicate is true and \texttt{t} otherwise.
This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
As with \texttt{or}, the variable must be local to the function
and it is not being assigned in another function.

@@

In our example, we are using the refinement of table types to
build the type of the variable \texttt{Shape},
as it should work as our class.
This means that the local \texttt{Shape} has type
\texttt{\{"x":number, "y":number\}} inside the definition of \texttt{new}.
This also makes Typed Lua bind the type of the local \texttt{Shape}
to the type \texttt{self} inside the definition of \texttt{new},
allowing us to access fields \texttt{x} and \texttt{y}.
After the definition of \texttt{new}, the local \texttt{Shape}
has type
\texttt{\{"x":number, "y":number, const "new":(self, number, number) -> (self)\}},
which is the type that Typed Lua binds to the type \texttt{self}
inside the definition of \texttt{move}.
We use the \texttt{const} annotation in the type of the methods
because it is necessary for covariance among object types work.
Finally, the top-level \texttt{return} statement fixes the type of
\texttt{Shape}, because \emph{fixed} table types do not allow width
subtyping, and this behavior allows Typed Lua to use the refinement
of table types to type check single inheritance, as we will see in
this section.

>>>>

Typed Lua is an optional type system for Lua, and its main goal is to
provide static type checking for Lua.
To do that, Typed Lua extends the syntax of Lua 5.3 to introduce
optional type annotations, and performs local type inference \cite{pierce2000lti}
to detect more precise types for unannotated expressions.
Even though the compiler warns the programmer about type errors,
it always removes the type annotations to generate Lua code that
runs in unmodified Lua implementations.

Another goal of Typed Lua is to be backwards compatible with Lua.
This means that any Lua code is valid Typed Lua code.
To be backwards compatible with Lua, the syntactic extensions introduced
by Typed Lua do not include new reserved words.
Appendix \ref{app:syntax} presents the complete syntax of Typed Lua
in extended BNF.

We use the consistent-subtyping relation of gradual typing
\cite{siek2007objects,siek2013mutable} to formalize Typed Lua,
though it does not insert run-time checks in the gradual typing style.
In gradual typing, run-time checks inspect the interaction between
dynamically typed and statically typed code to guarantee that dynamically
typed code does not violate statically typed code during run-time.
We did not insert run-time checks at this moment because they can decrease
run-time performance \cite{allende2013cis}.
We believe that a careful evaluation of run-time checks should be done before
inserting them in the type system.
However, this evaluation is out of scope of this work.

Unlike Dart \cite{dart} and TypeScript \cite{typescript},
we are designing Typed Lua aiming soundness to make it possible to
switch Typed Lua from optional typing to gradual typing in the future.
A sound type system is a prerequisite to insert run-time checks after
static type checking, because a sound type system ensures that
statically typed code will not throw type errors during run-time.

@@

Lua programmers often overload the input parameters of functions,
and use the \texttt{type} function to inspect the tag of the
input parameters to take different actions depending on what
those tags are.
The simplest case overloads on just a single parameter:
\begin{verbatim}
    local function overload (s1:string, s2:string|integer)
      if type(s2) == "string" then
        return s1 .. s2
      else
        -- string.rep : (string, integer, string?) -> (string)
        return string.rep(s1, s2)
      end
    end
\end{verbatim}

Typed Lua has a small set of \texttt{type} predicates that allows
programmers to constrain the type of a local variable inside a condition.
This example uses the predicate \texttt{type(v) == "string"} that
constrains the type of \texttt{v} from \texttt{string|t} to
\texttt{string} when the predicate is true and \texttt{t} otherwise.
This is a simplified form of \emph{flow typing} \cite{guha2011tlc,tobin-hochstadt2010ltu}.
As with \texttt{or}, the variable must be local to the function
and it is not being assigned in another function.

@@

In our example, we are using the refinement of table types to
build the type of the variable \texttt{Shape},
as it should work as our class.
This means that the local \texttt{Shape} has type
\texttt{\{"x":number, "y":number\}} inside the definition of \texttt{new}.
This also makes Typed Lua bind the type of the local \texttt{Shape}
to the type \texttt{self} inside the definition of \texttt{new},
allowing us to access fields \texttt{x} and \texttt{y}.
After the definition of \texttt{new}, the local \texttt{Shape}
has type
\texttt{\{"x":number, "y":number, const "new":(self, number, number) -> (self)\}},
which is the type that Typed Lua binds to the type \texttt{self}
inside the definition of \texttt{move}.
We use the \texttt{const} annotation in the type of the methods
because it is necessary for covariance among object types to work.
Finally, the top-level \texttt{return} statement fixes the type of
\texttt{Shape}, because \emph{fixed} table types do not allow width
subtyping, and this behavior allows Typed Lua to use the refinement
of table types to type check single inheritance, as we will see in
this section.
